### 2019-12-20 R/msBalance.Rmd
changed file extensions
more changes
Added bibliographic information, initial frequency counts.

---
title: "Balance responses in MS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{msBalance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

@@

```{r}
source("../paths.txt")
```

## Locate sources
Reading source files from `r dataDir`

## Run Matlab code to load data

>>>>

---
title: "Balance responses in MS"
author:
- J. Lucas McKay
- Daniel S. Peterson
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
bibliography: bibliography.bib
csl: nature.csl
---

@@

```{r}
source("../paths.m")
```

# Introduction
This document describes the analysis of standing balance responses in people with and without multiple sclerosis (MS) using wavelet-based functional ANOVA (wfANOVA[@mckay2013]).
MS has profound effects on balance,[@vanliew2019] including altering protective stepping behaviors.[@peterson2016]
The main objective of this study was to examine the impact of MS on the pattern of magnitude and timing of muscle activation during feet-in-place balance responses.

# Methods and Results

## Data sources
```{r, message=FALSE, warning=FALSE}
library(vroom)
library(dplyr)
library(stringr)
library(lubridate)
`%contains%` = function(s, pattern) grepl(pattern, s)
`%omits%` = function(s, pattern) !grepl(pattern, s)

# load participant-level information
m = vroom("../csv/m.csv") %>% mutate(Group = "MS") %>% mutate(Age = round(age)) %>% mutate(Sex = toupper(gender)) %>% rename(Participant = Subj)
c = vroom("../csv/c.csv") %>% mutate(Group = "Control") %>% mutate(Age = round(age)) %>% mutate(Sex = toupper(gender)) %>% rename(Participant = Subj)

# concatenate the participants. note that some of them appear to be coded as follow-ups. this is very useful, but may not be needed for the first report. they appear to be coded as such by the presence of "F1", "F2" in the participant name. convert this to a visit number (1, 2, 3, 4)
participants = full_join( m %>% select(-trial, -age, -gender) %>% distinct(), c %>% select(-trial, -age, -gender) %>% distinct() ) %>%
  mutate(`Visit Number` = ifelse(Participant %omits% "F\\d", 1, as.numeric(str_extract(str_extract(Participant, "F\\d"),"\\d"))+1)) %>% 
  mutate(`Visit Date` = date(dmy_hms(obsDate))) %>% select(-obsDate) %>% 
  mutate(`Visit ID` = Participant) %>% 
  mutate(`Visit Code` = str_extract(Participant, "F\\d")) %>% 
  mutate(Participant = str_replace(Participant, "F\\d", "")) %>% 
  mutate(Group = factor(Group, levels = c("MS","Control")))

# identify enrollment date for each participant and calculate amount of observation time
participants = participants %>% left_join(participants %>% group_by(Participant) %>% summarize(`Enrollment Date` = min(`Visit Date`, na.rm = TRUE))) %>% 
  mutate(`Time Since Enrollment (Days)` = ifelse(Group == "MS", `Visit Date` - `Enrollment Date`, 0))

# calculate amount of follow-up
followUp = participants %>% select(Group, Participant, `Time Since Enrollment (Days)`) %>% group_by(Group, Participant) %>% summarize(`Study Visits Available (n)` = n(), `Follow-Up Available (Days)` = max(`Time Since Enrollment (Days)`))

```

This study performed a secondary analysis on existing electromyographic recordings during balance responses in people with and without MS.
Individual subjects were as follows:

```{r, warning=FALSE, message=FALSE}
library(janitor)
library(kableExtra)
median0 = function (x) median(x[x>0], na.rm = TRUE)
min0 = function (x) min(x[x>0], na.rm = TRUE)
max0 = function (x) max(x[x>0], na.rm = TRUE)

k = function (x) kable_styling(x, full_width = F, position = "left")
t = followUp %>% tabyl(Group) %>% adorn_totals("row") %>% adorn_pct_formatting(digits = 0) %>% rename(N = n, `Percent`=percent)
followUpAvailable = followUp %>% filter(Group=="MS") %>% ungroup() %>% mutate(f = `Follow-Up Available (Days)`>0) %>% summarize(fAvail = sum(f==TRUE,na.rm = TRUE)/n()) %>% pull() %>% scales::percent()
followUpMed = median0(followUp$`Follow-Up Available (Days)`)
followUpMin = min0(followUp$`Follow-Up Available (Days)`)
followUpMax = max0(followUp$`Follow-Up Available (Days)`)
kable(t) %>% k()
```

Of these, follow-up visits were available for several of the participants with MS (≈`r followUpAvailable`).
Among the participants with MS and follow-up available, the median duration was `r followUpMed` days, and the range was from `r followUpMin` to `r followUpMax` days.
The number of visits was distributed as follows:

```{r, warning=FALSE, message=FALSE}
library(janitor)
library(kableExtra)
t = followUp %>% tabyl(Group, `Study Visits Available (n)`)
kable(t) %>% k() %>% add_header_above(c(" " = 1, "Number of Study Visits Available" = 4))
```

## Exclusions
Beginning with N=TK records, TK were excluded (N=TK, MS, N=TK, Control) due to incomplete demographic information.
Characteristics of the remaining participants included in analyses are summarized in Table 1.

## Demographic and Clinical Characteristics

# Disclosures

## Funding sources and conflict of interest
This study was supported in part by National Institutes of Health grant K25HD086276.
The authors report no conflict of interest related to this study.

## Financial disclosures for the previous 12 months
Dr. McKay has received research funding or support from the National Institutes of Health. Dr. McKay has received consulting fees from BioCircuit Technologies.

# References

### 2019-12-20 README.Rmd
updated path locations
more changes
Added bibliographic information, initial frequency counts.

1. Use RStudio to create a local copy of the project
    1. In RStudio, start a new Project: `File > New Project > Version Control > Git`
    1. In the “repository URL” paste: https://github.com/jlucasmckay/msBalance
    1. On OS X, you can make this under `tmp` - this is a temporary file that can be deleted later
    1. If there are files that need to be added to the remote, they can be placed in this folder or dragged / dropped into the GitHub interface.
1. Path locations (used by Matlab code) are in `msBalance/paths.m`
1. The main analysis is in `msBalance/R/msBalance.Rmd`
1. Stage changed files, commit, and push changes with the git panel.

@@

>>>>

1. Use RStudio to create a local copy of the project
    1. In RStudio, start a new Project: `File > New Project > Version Control > Git`
    1. In the “repository URL” paste: https://github.com/jlucasmckay/msBalance
    1. On OS X, you can make this under `tmp` - this is a temporary file that can be deleted later
    1. If there are files that need to be added to the remote, they can be placed in this folder or dragged / dropped into the GitHub interface.
1. Path locations that need to be read by Matlab and R can be specified in `msBalance/paths.m`.
1. Path locations might need to be modified in `msBalance/m/loadData.m`.
1. The main analysis is in `msBalance/R/msBalance.Rmd`
1. Stage changed files, commit, and push changes with the git panel.

@@

### Running the code

1. Alter `paths.m` to point to the correct data directory.
1. Run the Matlab code to load .mat files and populate patient-level data
    * Run `m/loadData.m` to load source files and populate the `csv/` directory.
    * A future version of this could call the Matlab script from R if useful.
1. "knit" `R/msBalance.Rmd` to produce the study document.
